### 2015-06-19 chapters/ipa_meets_unicode.tex
major update of IPA chapter

The International Phonetic Alphabet (IPA) is a common standard in linguistics to
transcribe sounds of spoken language into Latin-based characters
\citep{IPA2005}. Although IPA is reasonably easily adhered to with pen and
paper, it is not trivial to encode IPA characters electronically. Similar to the
previous chapter, this chapter will discuss various pitfalls with the encoding
of the IPA.\@ The details of the encoding are unimportant as long as the
transcription is only directed towards phonetically trained eyes. For a linguist
reading an IPA transcription, many of the details that will be discussed in this
chapter might seem like hair-splitting trivialities. However, if IPA
transcriptions are intended to be used across resources (e.g.~searching similar
phenomena across different languages) then it becomes crucial that there are strict
encoding guidelines. Our main goal in this chapter is to present the encoding
issues and propose recommendations for a ``strict'' IPA encoding.

@@

As explained in Section~\ref{pitfall-blocks}, the Unicode Standard code space is
subdivided into character blocks, which generally encode characters from a
single script. However, as is illustrated by the IPA, characters that form a
single writing system may be dispersed across several different character
blocks. With its diverse collection of symbols from various scripts and
diacritics, the IPA is spread across 12 blocks in the Unicode
Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned
symbols are counted or if the phonetic symbols commonly found in the literature
are also included, see~\cite[Appendix~C]{Moran2012}.}

\begin{itemize}
	\item Basic Latin (27 characters): \newline < a b c d e f h i j k l m n o p q r s t u v w x y z~.~| >
	\item Latin-1 Supplement (4 characters): \newline < æ ç ð ø >
	\item Latin Extended-A (3 characters): \newline < ħ ŋ œ >
	\item Latin Extended-B (4 characters): \newline < ǀ ǁ ǂ ǃ >
	\item IPA Extensions (67 characters): \newline < ɐ ɑ ɒ ɓ ɔ ɕ ɖ ɗ ɘ ə ɛ ɜ ɞ ɟ ɠ ɡ ɢ ɣ ɤ ɥ ɦ ɧ ɨ ɪ ɬ ɭ ɮ ɯ ɰ ɱ ɲ ɳ ɴ ɵ ɶ ɸ ɹ ɺ ɻ ɽ ɾ ʀ ʁ ʂ ʃ ʄ ʈ ʉ ʊ ʋ ʌ ʍ ʎ ʏ ʐ ʑ ʒ ʔ ʕ ʘ ʙ ʛ ʜ ʝ ʟ ʡ ʢ > 
	\item Spacing Modifier Letters (17 characters): \newline < ʰ ʲ ʷ ʼ ˈ ˌ ː ˑ ˞ ˠ ˡ ˤ ˥ ˦ ˧ ˨ ˩ >
	\item Combining Diacritical Marks (29 characters): \newline < {\large \  ̃\ \ ̆\ \ ̈\ ̚\ \ ̘\ \ ̙\ \ \ ̜\ \ ̝\ \ ̞\ \ ̟\ \ ̠\ \ ̤\ \ ̥\ \ ̩\ \ ̪\ \ ̬\ \ ̯\ \ ̰\ \ ̹\ \ ̺\ \ ̻\ \ ̼\ \ ̽\ \ ͜\  ̋\ \ ́\ \ ̄\ \ ̀\ \ ̏\ } >
	\item Greek and Coptic (3 characters): \newline < β θ χ >
%	\item Phonetic Extensions (2 characters): \charis{ᴅ ᴴ} 
%	\item Phonetic Extensions Supplement (1 characters): \newline < \charis{ᶑ} >
    \item General Punctuation (2 characters): \newline < ‖ \charis{‿} >
	\item Superscripts and Subscripts (1 character): \newline < ⁿ > 
	\item Arrows (4 characters): \newline < ↑ ↓ ↗ ↘ >
	\item Latin Extended-C (1 character): \newline < \charis{ⱱ} >
\end{itemize}

@@

Another problem is the large number of homoglyphs, i.e.~different characters
that have highly similar glyphs (or even completely identical glyphs, depending
on the font rendering). For example, a user of a Cyrillic computer keyboard
should ideally not use the <а> \textsc{cyrillic small letter a} at code point
\uni{0430} for IPA transcriptions, but instead use the <a> \textsc{latin small
letter a} at code point \uni{0061}, although visually they are mostly
indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic
keyboard. Some problematic homoglyphs related to encoding IPA in the Unicode Standard are 
the following.
\begin{itemize}

   \item The uses of the apostrophe has led to long discussions on the Unicode
        Standard email list. An English keyboard inputs <{\fontspec{Monaco}'}>
        \textsc{apostrophe} at \uni{0027}, although the preferred Unicode
        apostrophe is the <\ ' > \textsc{right single quotation mark} at
        \uni{2019}.\footnote{Note that many word processors (like Microsoft
        Word) by default will replace straight quotes by curly quotes,
        depending on the whitespace around it.} However, the glottal
        stop/glottalization/ejective marker is yet another completely different
        character <{\large ʼ}>, namely the \textsc{modifier letter apostrophe} at
        \uni{02BC}, which unfortunately looks mostly extremely similar to
        \uni{2019}. 
  \item Another problem is the <ˁ> \textsc{modifier letter reversed glottal
        stop} at \uni{02C1} vs.\@ the <ˤ> \textsc{modifier letter small reversed
        glottal stop} at \uni{02E4}. Both appear in various resources
        representing phonetic data online. This is thus a clear example for
        which the Unicode Standard does not solve the linguistic standardization
        problem.
  \item Even linguists are unlikely to distinguish between the <ə>
        \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
        \textsc{latin small letter turned e} at \uni{01DD}.
  \item The alveolar click <ǃ> at \uni{01C3} is of course often simply
        typed as <!> \textsc{exclamation mark} at \uni{0021}.\footnote{In the
        Unicode Standard the <ǃ> at \unif{01C3} is labeled \textsc{latin letter
        retroflex click}, but in IPA that glyph is used for an alveolar or
        postalveolar click (not retroflex). This naming is probably best seen as
        an error in the Unicode Standard.}
  \item The dental click <ǀ>, in Unicode known as \textsc{latin letter dental
        click} at \uni{01C0}, is of course often simply typed as <|>
        \textsc{vertical line} at \uni{007C}.
  \item For the marking of length there is a special Unicode characters, namely
        <ː> \textsc{modifier letter triangular colon} at \uni{02D0}. However,
        typing <:> \textsc{colon} at \uni{003A} is of course much easier.

@@

Making things even more problematic, there is an old typographic tradition that
the double-story <a> uses a single-story <ɑ> in italics. This leads to the
unfortunate effect that in most well-designed fonts the italics of <a> and <ɑ>
use the same glyph. If this distinction has to be kept upright in italics, the
only solution we can currently offer is to use \textsc{slanted} glyphs
(i.e.~artificially italicized glyphs) instead of real italics (i.e.~special
italics glyphs designed by a typographer).\footnote{For example, the widely used
IPA font Doulos SIL
(\url{http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL}) does not
have real italics. This leads some word-processing software, like Microsoft
Word, to produce slanted glyphs instead. That particular combination of font and
software application will thus lead to the desired effect distinguishing <a>
from <ɑ> in italics. However, note that when the text is transferred to another
font (i.e.~one that includes real italics) and/or to another software
application (like Apple Pages, which does not perform slanting), then this
visual appearance will be lost. In this case we are thus still in the
pre-Unicode situation in which the choice of font and rendering software
actually matters. The ideal solution from a linguistic point of view would be
the introduction of a new IPA code point for a different kind of <a>, which
explicitly specifies that it should still be rendered as a double-story
character when italicized. After informal discussion with various Unicode
players, our impression is that this highly restricted problem is not
sufficiently urgent to introduce even more <a> homographs in Unicode (which
already lead to much confusion, see Section~\ref{pitfall-homoglyphs}).}

@@

Reversely, there are a few cases in which the IPA distinguishes different
concepts, but the visual marking proposed by the IPA look very much alike. Such
cases are thus homoglyphs in the IPA itself, which of course need different
encodings.

@@

   \item The dental click <ǀ> and the indication of a minor group break <|>
           (proposed to be used for separating feet) look almost the same in
           most fonts. For a proper encoding, the \textsc{latin letter dental
           click} at \uni{01C0} and the \textsc{vertical line }at \uni{007C} 
           should be strictly distinguished.
   \item Similarly, the alveolar lateral click <ǁ>	should be encoded with a
           \textsc{latin letter lateral click} at \uni{01C1}, different from <‖>, 
           which according to the IPA is a major group break (by intonation), to be
           encoded by \textsc{double vertical line} at \uni{2016}.
   \item The marking of primary stress < ˈ > looks like an apostrophe, and
           is often typed with the same symbol as the ejective < ʼ >. For a
           proper encoding, these two symbols should be typed as 
           \textsc{modifier letter vertical line} at \uni{02C8} and
           \textsc{modifier letter apostrophe} at \uni{02BC}, respectively. 
   \item There are two different ``dashed''-l characters in IPA, namely the <ɫ> 	      \textsc{latin small letter l with middle tilde} at \uni{026B} and the <ɬ>	      \textsc{latin small letter l with belt} at \uni{026C}. These of course
          look highly similar, although they are different sounds. As a solution, 
          we will actually propose to not use the middle tilde at all 
          (see Section~\ref{pitfall-multiple-options-ipa}).

@@

% ==========================
\section{Pitfall: Multiple options in IPA}
\label{pitfall-multiple-options-ipa}     
% ==========================

It is not just the Unicode Standard that offers multiple options for encoding
the IPA.\@ Even the IPA specification itself offers some flexibility in how
transcriptions have to be encoded. There are a few cases in which the IPA
explicitly allows for different options of transcribing the same phonetic
content. This is understandable from a transcriber's point of view, but it is
not acceptable for interoperability between resources written in IPA.\@ We
consider it crucial to distinguish between ``lax'' IPA, for which it is
sufficient that any phonetically-trained reader is able to understand the
transcription, and ``strict'' IPA, which should be standardized on a single
unique encoding for each sound, so search will work across resources. We are 
aware of the following double options in the IPA, which will be discussed in 
turn below:

\begin{itemize}
  \item The marking of tone
  \item The marking of <g>
  \item The marking of velarization ad pharyngealization
  \item The placement of diacritics
\end{itemize}

The first case in which the IPA allows for different encodings is the question
of how to transcribe tone. There is an old tradition to use diacritics on vowels
to mark different tone levels, e.g. <ȅèée̋>.\footnote{To make things even more
complexe, there are at least two different Unicode homoglyphs for the low and
high level tones, namely <\ ̀~> \textsc{combining grave tone mark} at
\unif{0340} vs.\@ <\ ̀~> \textsc{combining grave accent} at \unif{0300} for low
tone, and <\ ́~> \textsc{combining acute tone mark} at \unif{0341} vs.\@ <\ ́~>
\textsc{combining acute accent} at \unif{0301} for high tone.} The IPA also
proposes the option of tone letters, e.g. <˥˦˧˨˩>, which are used much less, but
are more consistent for contours. Tone letters in the IPA have five different levels, and
sequences of these letters can be used to indicate contours. Well-designed fonts
will even merge a sequence of tone letters into a contour. For example, compare
the font Linux Libertine, which does not merge tone letters
<{\fontspec{LinLibertineO}˥˨˧˩}>, with the font CharisSIL, which merges this
sequence of four tone letters into a single contour <\charis{˥˨˧˩}>. For strict
IPA encoding we propose to standardize on tone letters.

@@

Third, the IPA has different markers for velarization <ˠ> and pharyngealization
<ˤ>. Confusingly, there is also a marker for ``velarized or pharyngealized'',
using the <\ \ {\large ̴}> \textsc{combining tilde overlay} at \uni{0334}. The
tilde overlay seems to be extremely rarely used. For strict IPA encoding we
propose to not include the combining tilde overlay at all, but only use the 
other two Spacing Modifiers.

Finally, for marking marking of voiceless pronunciation of voiced segments the IPA
uses the ring diacritic. Originally, the ring should be placed below the 
base character, like in <m̥>, using the \textsc{combining ring below} at \uni{0325}. 
However, in letters with long descenders the IPA also allows to put the ring 
above the base, like in <ŋ̊>, using the \textsc{combining ring above} at 
\uni{030A}. Yet, proper font design does not have any problem with rendering 
the ring below the base character, like in <ŋ̥>, so for strict IPA encoding we 
propose to standardize on the ring below. As a principle, for strict IPA 
encoding only one options should be allowed.

@@

The IPA includes a tie bar to indicate affricates, doubly articulated consonants
or diphthongs. This character can be placed above or below the base characters,
e.g. <\charis{t͡s}> or <\charis{t͜s}>. IPA allows both options. The choice
between the two symbols is purely for legible rendering; there is no difference
in semantics between the two symbols. The tie-bar is a convenient diacritic for
visually tokenizing input strings into chunks of phonetically salient groups.
However, rendering is such a problematic issue for tie bars in general that many
linguists simply do not use them. Just looking at a few different fonts already
indicates that actually no font designer really gets the placement right in
combination with superscripts and subscripts. If really necessary, we propose to
standardize on the tie bar above the base characters, using Unicode
\textsc{combining double inverted breve} at \uni{0361}.

\begin{itemize}
  \item[] {\fontspec{Times New Roman}Times new Roman: <t̥ʰ͡s> <t̥ʰ͜s>}
  \item[] {\small \fontspec{CharisSIL}CharisSIL:\@ <t̥ʰ͡s> <t̥ʰ͜s>}
  \item[] {\footnotesize \fontspec{Monaco}Monaco: <t̥ʰ͡s> <t̥ʰ͜s>}
  \item[] {\fontspec{DoulosSIL}DoulosSIL:\@ <t̥ʰ͡s> <t̥ʰ͜s>}
  \item[] Linux Libertine: <t̥ʰ͡s> <t̥ʰ͜s>
\end{itemize}

Tie bars are a special type of character in the sense that they do not belong to
a segment, but bind two graphemes together. This actually turns out to be rather
different from the Unicode conceptions, see
Section~\ref{pitfall-different-notions-of-diacritics}. According to the spirit
of the IPA, it is possible to combine more than two base characters into one tie
bar, but this is not possible with Unicode (i.e.~there is no possibility to draw
a tie bar over three of four characters). It is possible to indicate such larger
groups by repeating the tie bar, like in <a͡o͡u>. If really necessary, we
consider this possible, even though the rendering will never look good.

Most importantly though, the tie-bar is unnecessary for computational
tokenization given the design of orthography profiles, as described in
Section~\ref{usecase-tokenization}. The principle idea of such tokenization is 
that groups of characters that belong together are defined in an orthography 
profile, making it superfluous to indicate them explicitly by tie bars. For this 
reason we actually would like to suggest to remove the tie bar from any 
``strict'' IPA encoding.

@@

Although many combinations of base character with diacritic are treated as
canonical equivalent with precomposed characters, there are a few combinations
in IPA that allow for multiple, apparently identical, encodings that are not
canonical equivalent (see Section~\ref{pitfall-canonical-equivalence}),. The
following elements should not be treated as diacritics when encoding IPA in
Unicode: 
\begin{itemize}
  \item[] <\ {\large  ̡}\ > \textsc{combining palatalized hook below} at \uni{0321}
  \item[] <\ \ {\large  ̢}> \textsc{combining retroflex hook below} at \uni{0322}
  \item[] <\ \ {\large  ̵}> \textsc{combining short stroke overlay} at \uni{0335}
  \item[] <\ \ {\large  ̷}> \textsc{combining short solidus overlay} at \uni{0337}
  \item[] <\ \ {\large  ̴}> \textsc{combining tilde overlay} at \uni{0334}
\end{itemize}

There turn out to be a lot of characters in the IPA that could be conceived as
using any of these elements, like <ɲ>, <ɳ>, <ɨ>, <ø> or <ɫ>. However, all such
characters exist as well as precomposed combination in Unicode, and these
precomposed characters should preferably be used.\footnote{The IPA does not
describe any character for a voiced retroflex implosive, which would
transparently be \charis{ᶑ}. We propose to add this character to the IPA.} When
combinations of a base character with diacritic are used, then these
combinations are not canonical equivalent to the precomposed combinations. This
means that any search will not find both at the same time.

@@

Reversely, note that the <ç> \textsc{latin small letter c with cedilla} at
\uni{00E7} will be decomposed into <c> with <\ \ {\large ̧}> \textsc{combining
cedilla} at \uni{0327} by Unicode canonical decomposition, also if such a
decomposition is not intended in the IPA.\@

@@

	\item[] Length marks, namely: 
	\begin{itemize}
	  \item[] <ː> \textsc{modifier letter triangular colon} at \uni{02D0}
	  \item[] <ˑ> \textsc{modifier letter half triangular colon} at \uni{02D1}
	\end{itemize}

	\item[] Tone letters, like: 
	\begin{itemize} 
	  \item[] <˥> \textsc{modifier letter extra-high tone bar} at \uni{02E5}
	  \item[] <˧> \textsc{modifier letter mid tone bar} at \uni{02E7}
	  \item[] and others like this
	\end{itemize}

>>>>

The International Phonetic Alphabet (IPA) is a common standard in linguistics to
transcribe sounds of spoken language into Latin-based characters. Although IPA
is reasonably easily adhered to with pen and paper, it is not trivial to encode
IPA characters electronically. Similar to the previous chapter, this chapter
will discuss various pitfalls with the Unicode encoding of the IPA.\@ We will
specifically refer to the 2005 version of the IPA \citep{IPA2005} and the 7.0 
version of Unicode \citet{Unicode2014}.

The details of the encoding are unimportant as long as the
transcription is only directed towards phonetically trained eyes. For a linguist
reading an IPA transcription, many of the details that will be discussed in this
chapter might seem like hair-splitting trivialities. However, if IPA
transcriptions are intended to be used across resources (e.g.~searching similar
phenomena across different languages) then it becomes crucial that there are strict
encoding guidelines. Our main goal in this chapter is to present the encoding
issues and propose recommendations for a ``strict'' IPA encoding.

@@

As explained in Section~\ref{pitfall-blocks}, the Unicode Standard code space is
subdivided into character blocks, which generally encode characters from a
single script. However, as is illustrated by the IPA, characters that form a
single writing system may be dispersed across several different character
blocks. With its diverse collection of symbols from various scripts and
diacritics, the IPA is spread across 12 blocks in the Unicode
Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned
symbols are counted or if the phonetic symbols commonly found in the literature
are also included, see~\cite[Appendix~C]{Moran2012}. The 107 characters from 12 
code blocks shown here are the characters proposed for ``strict'' IPA encoding, 
as discussed in Section~\ref{ipa-recommendations}.}

\begin{itemize}[itemsep=4pt]
	\item \textsc{Basic Latin }(27 characters) \newline 
	a b c d e f h i j k l m n o p q r s t u v w x y z~.~|
	\item \textsc{Latin-1 Supplement} (4 characters) \newline 
	æ ç ð ø
	\item \textsc{Latin Extended-A} (3 characters) \newline 
	ħ ŋ œ
	\item \textsc{Latin Extended-B} (4 characters) \newline 
	ǀ ǁ ǂ ǃ
	\item \textsc{IPA Extensions} (67 characters) \newline 
	ɐ ɑ ɒ ɓ ɔ ɕ ɖ ɗ ɘ ə ɛ ɜ ɞ ɟ ɠ ɡ ɢ ɣ ɤ ɥ ɦ ɧ ɨ ɪ ɬ ɭ ɮ ɯ ɰ ɱ ɲ ɳ ɴ \newline
	ɵ ɶ ɸ ɹ ɺ ɻ ɽ ɾ ʀ ʁ ʂ ʃ ʄ ʈ ʉ ʊ ʋ ʌ ʍ ʎ ʏ ʐ ʑ ʒ ʔ ʕ ʘ ʙ ʛ ʜ ʝ ʟ ʡ ʢ 
	\item \textsc{Spacing Modifier Letters} (17 characters) \newline
	\dia{02DE} \dia{02E1} \dia{02B7} \dia{02B2} \dia{02E0} \dia{02E4} 
	\dia{02B0} \dia{02BC} \dia{02D0} \dia{02D1} ˥ ˦ ˧ ˨ ˩ {\large ˈ ˌ}
	\item \textsc{Combining Diacritical Marks} (23 characters) \newline 
	\dia{033C} \dia{032A} \dia{033B} \dia{033A} \dia{031F} \dia{0320} \dia{031D} 
	\dia{031E} \dia{0318} \dia{0319} \dia{031C} \dia{0339} \dia{032C} \dia{0325} 
	\dia{0330} \dia{0324} \dia{0329} \dia{032F} \dia{0303} \dia{0308} \dia{033D} 
	\dia{0306} \dia{031A} 
	\item \textsc{Greek and Coptic} (3 characters) \newline 
	β θ χ
    \item \textsc{General Punctuation} (2 characters) \newline 
    ‖ \charis{‿}
	\item \textsc{Superscripts and Subscripts} (1 character) \newline
	\dia{207F} 
	\item \textsc{Arrows} (4 characters) \newline 
	↑ ↓ ↗ ↘
	\item \textsc{Latin Extended-C} (1 character): \newline
	\charis{ⱱ}
\end{itemize}

@@

Another problem is the large number of homoglyphs, i.e.~different characters
that have highly similar glyphs (or even completely identical glyphs, depending
on the font rendering). For example, a user of a Cyrillic computer keyboard
should ideally not use the <а> \textsc{cyrillic small letter a} at code point
\uni{0430} for IPA transcriptions, but instead use the <a> \textsc{latin small
letter a} at code point \uni{0061}, although visually they are mostly
indistinguishable, and the Cyrillic character is more easily typed on a Cyrillic
keyboard. Some further problematic homoglyphs related to encoding IPA in the
Unicode Standard are the following.

  \item The uses of the apostrophe has led to long discussions on the Unicode
        Standard email list. An English keyboard inputs the symbol
        <\dia{0027}> \textsc{apostrophe} at \uni{0027}, although the preferred Unicode
        apostrophe is the <\dia{2019}> \textsc{right single quotation mark} at
        \uni{2019}.\footnote{Note that many word processors (like Microsoft
        Word) by default will replace straight quotes by curly quotes,
        depending on the whitespace around it.} However, the glottal
        stop/glottalization/ejective marker is yet another completely different
        character, namely <\dia{02BC}>, i.e.~the \textsc{modifier letter apostrophe} 
        at \uni{02BC}, which unfortunately looks mostly extremely similar to
        \uni{2019}. 
  \item Another problem is the <\dia{02C1}> \textsc{modifier letter reversed glottal
        stop} at \uni{02C1} vs.\@ the <\dia{02E4}> \textsc{modifier letter small reversed
        glottal stop} at \uni{02E4}. Both appear in various resources
        representing phonetic data online. This is thus a clear example for
        which the Unicode Standard does not solve the linguistic standardization
        problem.
  \item Linguists are also unlikely to distinguish between the <ə>
        \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
        \textsc{latin small letter turned e} at \uni{01DD}.
  \item The alveolar click <ǃ> at \uni{01C3} is of course often simply
        typed as <!> \textsc{exclamation mark} at \uni{0021}.\footnote{In the
        Unicode Standard the <ǃ> at \unif{01C3} is labeled \textsc{latin letter
        retroflex click}, but in IPA that glyph is used for an alveolar or
        postalveolar click (not retroflex). This naming is probably best seen as
        an error in the Unicode Standard. For the ``real'' retroflex click, see 
        Section~\ref{ipa-additions}.}
  \item The dental click <ǀ>, in Unicode known as \textsc{latin letter dental
        click} at \uni{01C0}, is often simply typed as <|> \textsc{vertical
        line} at \uni{007C}.
  \item For the marking of length there is a special Unicode characters, namely
        <\dia{02D0}> \textsc{modifier letter triangular colon} at \uni{02D0}. However,
        typing <\dia{003A}> \textsc{colon} at \uni{003A} is of course much easier.

@@

Making things even more problematic, there is an old typographic tradition that
the double-story <a> uses a single-story <ɑ> in italics. This leads to the
unfortunate effect that even in many well-designed fonts the italics of <a> and
<ɑ> use the same glyph. For example, in Linux Libertine (the font of this book)
the italics of these characters are highly similar <\textit{a}> and
<\textit{ɑ}>, while in Charis SIL they are identical: <\textit{\charis{a}}> and
<\textit{\charis{ɑ}}>. If this distinction has to be kept upright in italics,
the only solution we can currently offer is to use \textsc{slanted} glyphs
(i.e.~artificially italicized glyphs) instead of real italics (i.e.~special
italics glyphs designed by a typographer).\footnote{For example, the widely used
IPA font Doulos SIL
(\url{http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL}) does not
have real italics. This leads some word-processing software, like Microsoft
Word, to produce slanted glyphs instead. That particular combination of font and
software application will thus lead to the desired effect distinguishing <a>
from <ɑ> in italics. However, note that when the text is transferred to another
font (i.e.~one that includes real italics) and/or to another software
application (like Apple Pages, which does not perform slanting), then this
visual appearance will be lost. In this case we are thus still in the
pre-Unicode situation in which the choice of font and rendering software
actually matters. The ideal solution from a linguistic point of view would be
the introduction of a new IPA code point for a different kind of <a>, which
explicitly specifies that it should still be rendered as a double-story
character when italicized. After informal discussion with various Unicode
players, our impression is that this highly restricted problem is not
sufficiently urgent to introduce even more <a> homographs in Unicode (which
already lead to much confusion, see Section~\ref{pitfall-homoglyphs}).}

@@

Reversely, there are a few cases in which the IPA distinguishes different
phonetic concepts, but the visual characters used by the IPA look very much
alike. Such cases are thus homoglyphs in the IPA itself, which of course need
different encodings.

@@

   \item The dental click <ǀ> and the indication of a minor group break <|>
           look almost the same in
           most fonts. For a proper encoding, the \textsc{latin letter dental
           click} at \uni{01C0} and the \textsc{vertical line }at \uni{007C} 
           should be used, respectively.
   \item Similarly, the alveolar lateral click <ǁ>	should be encoded with a
           \textsc{latin letter lateral click} at \uni{01C1}, different from <‖>, 
           which according to the IPA is the character to by used for a major group 
           break (by intonation), to be encoded by \textsc{double vertical line} 
           at \uni{2016}.
   \item The marking of primary stress < ˈ > looks like an apostrophe, and
           is often typed with the same symbol as the ejective <\dia{02BC}>. For a
           proper encoding, these two symbols should be typed as 
           \textsc{modifier letter vertical line} at \uni{02C8} and
           \textsc{modifier letter apostrophe} at \uni{02BC}, respectively. 
   \item There are two different ``dashed''-l characters in IPA, namely the <ɫ> 	      \textsc{latin small letter l with middle tilde} at \uni{026B} and the <ɬ>	      \textsc{latin small letter l with belt} at \uni{026C}. These of course
          look highly similar, although they are different sounds. As a solution, 
          we will actually propose to not use the middle tilde at all 
          (see Section~\ref{pitfall-multiple-options-ipa}).

@@

% ==========================
\section{Pitfall: Multiple encoding options in IPA}
\label{pitfall-multiple-options-ipa}     
% ==========================

It is not just the Unicode Standard that offers multiple options for encoding
the IPA.\@ Even the IPA specification itself offers some flexibility in how
transcriptions have to be encoded. There are a few cases in which the IPA
explicitly allows for different options of transcribing the same phonetic
content. This is understandable from a transcriber's point of view, but it is
not acceptable when the goal is interoperability between resources written in
IPA.\@ We consider it crucial to distinguish between ``lax'' IPA, for which it
is sufficient that any phonetically-trained reader is able to understand the
transcription, and ``strict'' IPA, which should be standardized on a single
unique encoding for each sound, so search will work across resources. We are
aware of the following non-unique encoding options in the IPA, which will be
discussed in turn below:

\begin{itemize}
  \item The marking of tone
  \item The marking of <g>
  \item The marking of velarization and pharyngealization
  \item The placement of diacritics
\end{itemize}

The first case in which the IPA allows for different encodings is the question
of how to transcribe tone. There is an old tradition to use diacritics on vowels
to mark different tone levels, e.g. <ȅèée̋>.\footnote{To make things even more
complexe, there are at least two different Unicode homoglyphs for the low and
high level tones, namely <\diaf{0340}> \textsc{combining grave tone mark} at
\unif{0340} vs.~<\diaf{0300}> \textsc{combining grave accent} at \unif{0300} for
low tone, and <\diaf{0341}> \textsc{combining acute tone mark} at \unif{0341}
vs.<\diaf{0301}> \textsc{combining acute accent} at \unif{0301} for high tone.}
The IPA also proposes the option of tone letters, e.g. <˥˦˧˨˩>, which are much
less often used, but are more consistent for contours. Tone letters in the IPA
have five different levels, and sequences of these letters can be used to
indicate contours. Well-designed fonts will even merge a sequence of tone
letters into a contour. For example, compare the font Linux Libertine, which
does not merge tone letters <{\fontspec{LinLibertineO}˥˨˧˩}>, with the font
CharisSIL, which merges this sequence of four tone letters into a single contour
<\charis{˥˨˧˩}>. For strict IPA encoding we propose to standardize on tone
letters.

@@

Third, the IPA has special markers for velarization <\dia{02E0}> and
pharyngealization <\dia{02E4}>. Confusingly, there is also a marker for
``velarized or pharyngealized'', using the <\dia{0334}> \textsc{combining tilde
overlay} at \uni{0334}. The tilde overlay seems to be extremely rarely used. For
strict IPA encoding we propose to not include the combining tilde overlay at
all, but only use the other two special markers for velarization and
phayngaealization.

Finally, the IPA states that ``diacritics may be placed above a symbol with a
descender''. For example, for marking marking of voiceless pronunciation of
voiced segments the IPA uses the ring diacritic. Originally, the ring should be
placed below the base character, like in <m̥>, using the \textsc{combining ring
below} at \uni{0325}. However, in letters with long descenders the IPA also
allows to put the ring above the base, like in <ŋ̊>, using the \textsc{combining
ring above} at \uni{030A}. Yet, proper font design does not have any problem
with rendering the ring below the base character, like in <ŋ̥>, so for strict
IPA encoding we propose to standardize on the ring below. As a principle, for
strict IPA encoding only one options should be allowed for all diacritics.

The variable encoding as allowed by the IPA becomes even more troublesome for
the tilde and diaeresis diacritics. In these cases, the IPA itself attaches
different semantics to the symbols above and below a base characters. The tilde
above a character (like in <ã>, using the \textsc{combining tilde} at
\uni{0303}) is used for nasalisation, while the tilde below a character (like in
<a̰>, using the \textsc{combining tilde below} at \uni{0330}) indicates creaky
voice. Likewise, the diaeresis above (like in <ä>, using the \textsc{combining
diaeresis} at \uni{0308}) is used for centralization, while the diaeresis below
a character (like in <a̤>, using the \textsc{combining diaeresis below} at
\uni{0324}) indicates breathy voice. These cases strengthen our plea to not
allow diacritics to switch position for typographic convenience.

@@

The IPA includes a tie bar. The tie-bar is a convenient diacritic for visually
tokenizing input strings into chunks of phonetically salient groups, for example
to indicate affricates, doubly articulated consonants or diphthongs. This
character can be placed above or below the base characters, e.g. <\charis{t͡s}>
or <\charis{t͜s}>. IPA allows both options. The choice between the two symbols
is purely for legible rendering; there is no difference in semantics between the
two symbols. However, rendering is such a problematic issue for tie bars in
general that many linguists simply do not use them. Just looking at a few
different fonts already indicates that actually no font designer really gets the
placement right in combination with superscripts and subscripts. If really
necessary, we propose to standardize on the tie bar above the base characters,
using a \textsc{combining double inverted breve} at \uni{0361}.

\begin{itemize}[itemsep=6pt]
  \item[] {\fontspec{Times New Roman}Times new Roman: t̥ʰ͡s t̥ʰ͜s}
  \item[] {\small \fontspec{CharisSIL}CharisSIL:\@ t̥ʰ͡s t̥ʰ͜s}
  \item[] {\footnotesize \fontspec{Monaco}Monaco: t̥ʰ͡s t̥ʰ͜s}
  \item[] {\fontspec{DoulosSIL}DoulosSIL:\@ t̥ʰ͡s t̥ʰ͜s}
  \item[] Linux Libertine: t̥ʰ͡s t̥ʰ͜s
\end{itemize}

Tie bars are a special type of character in the sense that they do not belong to
a segment, but bind two graphemes together. This actually turns out to be rather
different from the Unicode conceptions, see
Section~\ref{pitfall-different-notions-of-diacritics}. Further, according to the
spirit of the IPA, it would also be possible to combine more than two base
characters into one tie bar, but this is not possible with Unicode (i.e.~there
is no possibility to draw a tie bar over three of four characters). It is
possible to indicate such larger groups by repeating the tie bar, like in
<a͡o͡u>. If really necessary, we consider this possible, even though the
rendering will never look good.

Most importantly though, the tie-bar is unnecessary for computational processing
given the design of orthography profiles, as described in
Section~\ref{usecase-tokenization}. The principle idea of such processing is
that groups of characters that belong together are defined in an orthography
profile, making it superfluous to indicate them explicitly by tie bars. For this
reason we suggest to remove the tie bar from any ``strict'' IPA encoding, and
instead add descriptions of all so-called ``tailored grapheme clusters'' (see
Section~\ref{the-unicode-approach}) to an orthography profile.

@@

Although many combinations of base character with diacritic are treated as
canonical equivalent with precomposed characters, there are a few combinations
in IPA that allow for multiple, apparently identical, encodings that are not
canonical equivalent (see Section~\ref{pitfall-canonical-equivalence}),. The
following elements should not be treated as diacritics when encoding IPA in
Unicode: 
\begin{itemize}
  \item[] <{\fontspec{CharisSIL}{\large ◌}}\symbol{"0321}> \textsc{combining palatalized hook below} at \uni{0321}
  \item[] <{\fontspec{CharisSIL}{\large ◌}}\symbol{"0322}> \textsc{combining retroflex hook below} at \uni{0322}
  \item[] <\dia{0335}> \textsc{combining short stroke overlay} at \uni{0335}
  \item[] <\dia{0337}> \textsc{combining short solidus overlay} at \uni{0337}
  \item[] <\dia{0334}> \textsc{combining tilde overlay} at \uni{0334}
\end{itemize}

There turn out to be a lot of characters in the IPA that could be conceived as
using any of these elements, like <ɲ>, <ɳ>, <ɨ>, <ø> or <ɫ>. However, all such
characters exist as well as precomposed combination in Unicode, and these
precomposed characters should preferably be used.\footnote{The IPA does not
describe any character for a voiced retroflex implosive, which would
transparently be \charis{ᶑ}. We propose to add this character to the IPA, see
Section~\ref{ipa-additions}.} When instead combinations of a base character with
diacritic are used, then these combinations are not canonical equivalent to the
precomposed combinations. This means that any search will not find both at the
same time.

@@

Reversely, note that the <ç> \textsc{latin small letter c with cedilla} at
\uni{00E7} will be decomposed into <c> with <\dia{0327}> \textsc{combining
cedilla} at \uni{0327} by Unicode canonical decomposition, also if such a
decomposition is not intended in the IPA.\@

@@

	\item[] Length marks, namely: 
	\begin{itemize}
	  \item[] <\dia{02D0}> \textsc{modifier letter triangular colon} at \uni{02D0}
	  \item[] <\dia{02D1}> \textsc{modifier letter half triangular colon} at \uni{02D1}
	\end{itemize}

	\item[] Tone letters, like: 
	\begin{itemize} 
	  \item[] <\dia{02E5}> \textsc{modifier letter extra-high tone bar} at \uni{02E5}
	  \item[] <\dia{02E8}> \textsc{modifier letter low tone bar} at \uni{02E8}
	  \item[] and others like this
	\end{itemize}
